RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-10-23 doc/paper.tex
Revizije u Sekciji 3.3.2

\subsubsection{Fitness function}\label{sec:fitness}

@@

Perturbations allow the algorithm to escape from local optima. As previously described, they are performed in two occasions: the first one during the exhaustive examination of neighboring solutions around the starting solution, and the second one being non-exhaustive, during selection of the next starting solution. 
In both cases, the same Algorithm~\ref{alg:pert} is used.

@@

This is followed by constant normalization during which coefficients that enter multiplication, division, addition or subtraction are set to 1, while those entering the power function are rounded to integer, with exception of square root, which is kept intact. For example, for expression $3.3\cdot(x+45.1\cdot y^{3.2})\cdot 81\cdot x/\sqrt{y}$ the normalized version is $1\cdot (x+1\cdot y^3)\cdot 1\cdot x/\sqrt{y}$. The reason for performing normalization is to reduce the search space of possible perturbations. This reduction is reasonable, since normalization preserves the essential functional form. (Note that coefficients get tuned later during the OLS phase and local search.)

After normalization, the expression is simplified -- getting compact expression is more likely after normalization than before. The previous expression will take form $(x+y^3)\cdot x/\sqrt{y}$ -- this simplification only removed unnecessary coefficients, but in general it can be non-trivial symbolic simplification.

@@

\emph{Case 1}. 
Subtree $ss$ is the whole tree $s$. Following on the the previous exemplary expression tree, this means that multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be changed to addition, which forms a perturbation $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Subtree $ss$ has arity of at least 1. This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $ss=x+y^3$ the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is changed to division). 
Another example would be the case of unary operation, e.g. when $ss=\sqrt{y}$. In that case, some of the possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

\emph{Case 3}. 
Subtree $ss$ is binary operation, meaning the right subtree must exist. 
The idea is analog as in \emph{Case 2}.

Allowed per-node changes are as follows:

\begin{enumerate}
	\item Any node to any of its subtrees (excluding itself). For example, if $(x+y^3)$ is changed to $x$, the perturbation is $(x+y^3)\cdot x/\sqrt{y} \rightarrow x\cdot x/\sqrt{y}$. 
	\item Variable to another variable or constant to variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (y+y^3)\cdot x/\sqrt{y}$.
	\item Variable to unary operation applied to that variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot \ln{x}/\sqrt{y}$.
	\item Unary operation to another unary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sin{y}$.
	\item Binary operation to another binary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot (x + \sqrt{y})$. 
	\item Variable or constant enter the binary operation with arbitrary variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+x/y^3)\cdot x/\sqrt{y}$. 
\end{enumerate}

>>>>

\fxnote{Dati neku smislenu procjenu kompleksnosti algoritma, ako to nije pretesko. }
\subsubsection{Fitness function}\label{sec:fitness}

@@

Perturbations allow the algorithm to escape from local optima. As previously described, perturbations are performed in two occasions: ($i$) during the exhaustive examination of neighboring solutions around the starting solution, ($ii$)during selection of the next starting solution, a non-exhaustive case.  
In both cases, the same Algorithm~\ref{alg:pert} is used.

@@

This is followed by constant normalization during which coefficients that enter multiplication, division, addition or subtraction are set to 1, while those entering the power function are rounded to integer, with exception of square root, which is kept intact. For example, for expression $3.3\cdot(x+45.1\cdot y^{3.2})\cdot 81\cdot x/\sqrt{y}$ the normalized version is $1\cdot (x+1\cdot y^3)\cdot 1\cdot x/\sqrt{y}$. The reason for performing normalization is reducing the search space of possible perturbations thus keeping them within a reasonable size. This reduction is reasonable, since normalization preserves the essential functional form. (Note that coefficients get tuned later during the OLS phase and local search.)

After performing the normalization process, the expression is simplified -- getting compact expression is more likely after normalization than before. The previous expression will take form $(x+y^3)\cdot x/\sqrt{y}$. The purpose of simplification is removing unnecessary coefficients, but in general it can also be non-trivial symbolic simplification.

@@

\emph{Case 1}. 
Subtree $ss$ is the whole tree $s$. Following on the the previous exemplary expression tree, this means that multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be replaced by addition, which forms a perturbation $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Subtree $ss$ has arity of at least 1. This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $ss=x+y^3$ the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is replaced by division). 
Another example would be the case of unary operation, e.g. when $ss=\sqrt{y}$. In that case, some of the possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

\emph{Case 3}. 
Subtree $ss$ is binary operation, meaning the right subtree must exist. 
The analogous idea is applied as in \emph{Case 2}.

The algorithm allows the following set of carefully chosen per-node changes:

\begin{enumerate}
	\item Any node to any of its subtrees (excluding itself). For example, if $(x+y^3)$ is changed to $x$, the perturbation is $(x+y^3)\cdot x/\sqrt{y} \rightarrow x\cdot x/\sqrt{y}$. 
	\item Variable to another variable or constant to variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (y+y^3)\cdot x/\sqrt{y}$.
	\item Variable to unary operation applied to that variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot \ln{x}/\sqrt{y}$.
	\item Unary operation to another unary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sin{y}$.
	\item Binary operation to another binary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot (x + \sqrt{y})$. 
	\item Variable or constant enter the binary operation with arbitrary variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+x/y^3)\cdot x/\sqrt{y}$. 
\end{enumerate} 
\fxnote{Dati kompleksnost velicine skupa perturbacija u odnosu na velicinu drveta i dopustene promjene}
